\newpage
\section*{Response to the Editor}
Dear Diego,

Thank you for giving us the opportunity to revise our paper and for your and the referees' constructive comments. We believe the revised version has significantly improved and we hope you will find it acceptable for publication. Below we summarize the major changes, followed by a point-by-point reply to referees.

\paragraph{Cities vs countries.} In our previous version, our analysis has jumped back and forth between cities and countries in a confusing manner. Both referees pointed out the restrictive assumptions needed to study cities in isolation. 

To maintain the analytical tractability of our model, we have kept the baseline analysis at the city level. We have correspondingly reworded our Introduction (pages XX) to avoid overselling our results. We also added a new Section XX, which explicitly aggregates up the city-level results to the country level. Our language is still cautious here, as our baseline model has little to say about the system of cities within a country. [We added Appendix XX to illustrate how a richer model can be used to study multiple cities within the same country.]

\paragraph{Better connection to literature.} We acknowledge the importance of earlier work in this literature and have rewritten the Introduction (pages XX) to carefully delineate our contribution. Indeed, the importance of land use, land restrictions and transport infrastructure for development has been highlighted by XX and XX. Our contribution is to study how the endogenous spatial sorting of sectors affects aggregate productivity.

\newpage
\section*{Response to Referee 1}

\newpage
\section*{Response to Referee 2}

